Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-18 manuscript/HIAs.qmd
select relevant NTs

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
---

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023], and web site).
This study builds on the work in that report.

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing (REF).
We included data fram 2018 to 2023.
In this survey, selected nature types are delineatied on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

-   Northern omgrotrophic \> 10.000 m^2^

-   All semi-natural mires (minerotrophic)

@@

-   Calcareous northen fens \>1000 m^2^

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones.
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

@@

![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay takingthe mean value).](../images/workflow.jpg){#fig-workflow}

@@

To

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
---

```{r setup, warning=F, message=F, include=F}
library(tidyverse)
library(knitr)
library(sf)

```

@@

In this study we focused on open mire ecosystems in Norway.
For this ecosystem we have an ecosystem delineation map produced using remotely sensed data and a deep learning model [@bakkestuen_delineation_2023].
This model, which has a 90.9% precision when tested against independent field data, estimates 12.7% of the area in southern Norway is mire [@bakkestuen_delineation_2023].
Mires are ecologically and socially important in Norway simply doe to its large extent, and due to its role in climate mitigation, as mires store a large amount of carbon (REF?).
There has not been a national assessment of the ecosystem condition of mires in Norway, but we recent contributed on a report which presented several new indicators that can be used in future assessment [@nybo_indikatorer_2023; see also @kolstad2023]).
This study builds on the work in that report.

@@

To describe these two concepts we use five condition variables from the same data set ( @tbl-variables).
The data set is from a standardised field survey of nature types in Norway that started in 2018 and which is still ongoing.
The data is freely available [@norwegian_environmental_agency_naturtyper_2024].
See @miljødirektoratet for the sampling protocol (in Norwegian).
We included data from 2018 to 2023.
In this survey, selected nature types are delineated on a map, and each locality is scored on a range of variables relevant for describing the state and quality of nature.
The surveys are commissioned with the goal of producing data relevant for immediate land-use decisions, and is therefore spatially biased, typically towards areas with high human impact or expected impact.
In addition there is a thematic and size bias in the sampling protocol.
Of all possible mire types, the survey only maps the following:

@@

-   Northern ombrotrophic mires \> 10.000 m^2^

-   All semi-natural mires (minerotrophic fens)

@@

-   Calcareous northern fens \>1000 m^2^

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones @asbjørnmoen1998 .
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed, and the variable 7FK is not recorded in ombrotrophic mires.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway.
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed we do not think this is the case for our variables at least.
It may also be that alien plants are more or less common on bogs relative to fens.
For this we have no assumption.

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|--------------|--------------|--------------|----------------|--------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

@@

![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}

@@

```{r filepath}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")

#pData <- ifelse(dir == "C:", 
#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
```

```{r}
# I already did some work to identify the relevant nature types

# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
naturetypes_summary <- readRDS("../data/natureType_summary.rds")

# Getting the nature types that have 7FA
myVars <- "7FA"
naturetypes_summary2 <- naturetypes_summary %>%
  rowwise() %>%
  mutate(keepers = sum(c_across(
    all_of(myVars))>0, na.rm=T)) %>%
  filter(keepers >0,
         Ecosystem == "våtmark") %>%
  select(Nature_type, NiN_mainType, Year)

naturetypes_summary2$Nature_type

# Getting the nature types that have the ADSV variables
myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
naturetypes_summary3 <- naturetypes_summary %>%
  rowwise() %>%
  mutate(keepers = sum(c_across(
    all_of(myVars))>0, na.rm=T)) %>%
  filter(keepers >0,
         Ecosystem == "våtmark") %>%
  select(Nature_type, NiN_mainType, Year)

naturetypes_summary3$Nature_type
```

```{r relevantNTs}
# These are the nature types I'm keeping for 7FA.
# Note the absence of ombrotriphic types
alienNTS <- c("Slåttemyr",
              "Semi-naturlig våteng",
              "Rik åpen jordvannsmyr i mellomboreal sone",
              "Rik åpen sørlig jordvannsmyr",
              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone")

# And these are the nature types that we use for the ADSV indicator
 advsNTs <- c(# fens
              "Slåttemyr",
              "Sørlig slåttemyr",
              "Rik åpen jordvannsmyr i mellomboreal sone",
              "Rik åpen sørlig jordvannsmyr",
              "Semi-naturlig myr",
              "Rik åpen jordvannsmyr i nordboreal og lavalpin sone",
              #"Sørlig kaldkilde", # excluded springs

              # bogs (mainly)
              "Høgereligende og nordlig nedbørsmyr",
              "Øyblandingsmyr",
              "Konsentrisk høymyr",
              "Sørlig nedbørsmyr",
              "Atlantisk høymyr",
              "Terrengdekkende myr",
              "Eksentrisk høymyr",
              "Kanthøymyr",
              "Platåhøymyr",
              "Palsmyr")

```

```{r muniMaps, include=F, eval=F}
naturetypes <- sf::st_read(dsn = path) %>%
  # keep only wetlands
  filter(hoved.kosystem == "våtmark") %>%

  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = case_match(hovedøkosystem,
                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                  .default = hovedøkosystem),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)

temp <- naturetypes %>%
    filter(hoved.kosystem == "våtmark")

```
